Enabling GPU Virtualization in Cloud Environments

### 2016-02-09 paper.tex
8 paginas

\usepackage{url}% http://ctan.org/pkg/url
\usepackage{flushend}

@@

\textcolor{black}{
Unfortunately, sharing GPU resources among multiple VMs in cloud environments is more complex than in physical servers.
On one hand, instances in public clouds are not easily customizable. 
On the other, although the instances in a private cloud can be customized in many aspects, when referring to GPUs
the number of options is reduced.
As a result, neither vendors nor tools offer GPGPU services.
}
Remote virtualization has been recently proposed to deal with the low-usage problem. Some relevant examples include {rCUDA}~\cite{tonithesis},
DS-CUDA~\cite{dscuda}, gVirtus~\cite{gvirtus}, vCUDA~\cite{vcuda}, VOCL~\cite{vocl}, and SnuCL~\cite{snucl}.

Roughly speaking these virtualization frameworks 
enable cluster configurations with fewer GPUs than nodes.  The goal is that 
GPU-equipped nodes act as GPGPU servers, yielding a GPU-sharing solution that potentially achieves
a higher overall utilization of the accelerators in the system.

@@

The rest of the paper is structured as follows. 
In Section~\ref{sec:background} we introduce the technologies used in this work;
Section~\ref{sec:state} summarizes related work; 
Section~\ref{sec:motivation} states the motivation to address this task;
the effort to use AWS is explained in Section~\ref{sec:workingaws};
while the work with Openstack is described in Section~\ref{sec:gpgpuOS};
finally, Section~\ref{sec:conclusions} summarizes the advances and Section~\ref{sec:future} outlines the next steps
 of this research.

\section{\uppercase{Background}}
\label{sec:background}
\subsection{The rCUDA Framework}
\label{sec:rcuda}
{rCUDA}~\cite{toniparco} is a middleware that enables transparent access
to any NVIDIA GPU device present in a cluster from all compute
nodes. The GPUs can also be shared among nodes, and a single node can use all the graphic accelerators
as if they were local.
These features aim to attain higher accelerator utilization rates in the overall system while simultaneously reducing
resource, space, and energy requirements~\cite{energy14}.
rCUDA is structured following a client-server distributed
architecture: the client middleware runs in the same cluster node where the application demanding GPGPU
acceleration services is executed, providing a transparent replacement for the
native CUDA libraries. Furthermore, the server middleware is executed in the
cluster nodes from which the actual GPUs provide the requested GPGPU services.
%To support a concurrent scenario where GPUs are shared between
%processes\slash nodes, {rCUDA} manages separate device contexts for
%each client application.

The {rCUDA} 5.0 client exposes the same interface as the regular NVIDIA
CUDA 6.5 release~\cite{cuda65}.%, including the runtime and driver
%APIs (Application Programming Interface) as well as other frequently used libraries such as cuBLAS, cuFFT, cuSparse or cuRand.
With this middleware, applications are not aware that they are executed
on top of a virtualization layer.
To deal with new GPU programming models, {rCUDA} has been recently extended to accommodate 
directive-based models such as OmpSs~\cite{repara15} and OpenACC~\cite{cluster15}.

The integration of remote GPGPU virtualization with global
resource schedulers such as SLURM~\cite{sbacpad14} completes this appealing 
technology, making accelerator-enabled clusters more flexible and
energy-efficient.

@@

OpenStack~\cite{OpenStack} is a cloud operating system (OS) that provides Infrastructure as a Service (IaaS). 
OpenStack controls large pools of compute, storage, and networking resources throughout a 
datacenter. All these resources are managed through a dashboard or an API that gives administrators 
control while empowering their users to provision resources through a web interface
or a command-line interface.  
OpenStack supports most recent hypervisors and handles provisioning 
and life-cycle management of VMs.

The OpenStack architecture offers flexibility to create a custom cloud, with no proprietary hardware
or software requirements, and the ability to integrate with legacy systems and third party technologies. 
\textcolor{black}{
From the HPC perspective, OpenStack offers high performance virtual machine configurations with different hardware architectures. 
Even though in OpenStack it is possible to work with GPUs, the Nova project does not support this architecture yet~\cite{OpenStackGPU}. 
}

@@

\section{\uppercase{GPGPU share service with AWS}}
\label{sec:workingaws}
In this section, we briefly expose the current user-level features 
provided by AWS, and then analyze this service  
from the point of view of GPU usage. 
Next, we describe the scenarios and experiments where virtualized GPUs are utilized and show a number of initial results.

\subsection{Current Features}
This subsection identifies several common features checked during 
the VM's creation step.

@@

An instance is a pre-configured VM focused on an 
specific target. Among the large list of instances offered by AWS, 
we can find specialized versions for general-purpose (T2, M4 and M3); computer science (C4 and C3); memory (R3); storage (I2 and D2) and 
GPU capable (G2). Each type of instance has its own purpose and cost (price). 
Moreover, each type offers a different number of CPUs as well as network 
interconnection, which can be: low, medium, high or 10Gb.

For our study, we worked in the AWS availability zone US EAST (N. VIRGINIA). 
The instances available in that case present the features reported in Table~\ref{table:awsInstances}.

@@

\subsection{Testbed Scenarios}
In this section, we introduce different client-server configurations. 
All scenarios are based on the maximum number of instances 
that a user can freely select without submitting a formal request. 
In particular, the maximum number for ``g2.2xlarge'' is 5; for ``g2.8xlarge'' it is 2.

All instances operate the RHEL 7.1 64-bit OS and version 6.5 of CUDA. 
We design three configuration scenarios for our tests:

@@

\begin{figure*}[ht]
\centering
\subfigure[5 remote GPUs over 1GbE.]{%
\includegraphics[width=0.315\linewidth]{images/aws3.pdf}
\label{subfig:aws1}}
\quad
\subfigure[8 remote GPUs over 10GbE.]{%
\includegraphics[width=0.315\linewidth]{images/aws2.pdf}
\label{subfig:aws2}}
\subfigure[13 remote GPUs over 1GbE.]{%
\includegraphics[width=0.315\linewidth]{images/aws1.pdf}
\label{subfig:aws3}}
\caption{Configurations where an AWS instance, acting as a client, employs several remote GPUs.}
\label{fig:aws}
\end{figure*}

@@

\subsection{Experimental results}
In this subsection we employ two GPU applications to assess the benefits of GPU virtualization.

@@

\subsection{Discussion}
\textcolor{black}{The previous experiments reveal that the AWS GPU-instances are not appropriate for HPC
because neither the network nor the accelerators are powerful enough to deliver high performance
when running compute-intensive parallel applications. As~\cite{tonithesis} demonstrates, network and device types are critical factors to performance.
In other words, AWS is more oriented toward offering a general-purpose service than to provide high performance.
Also, AWS fails in offering flexibility as it enforces the user to choose between a basic instance with a GPU and a powerful instance with 4 GPUs.}
Table~\ref{table:awsInstances} shows that the resources of the ``g2.8xlarge'' are quadrupled, but so is the cost per hour.
Therefore, in the case of having other necessities (instances type), using GPU virtualization technology we could in principle attach an accelerator to any type of instance.
Furthermore, reducing the budget spent in cloud services is possible by customizing the resources of the available instances.
For example, we can work on an instance with 2 GPUs for only \$ 1.3 by launching 2 ``g2.2xlarge'' and using remote GPUs, avoiding to pay the double for features that we do not need in ``g2.8xlarge''.
\textcolor{black}{
In terms of GPU-shareability, AWS reserves GPU-capable physical machines which will be waiting for a GPU-instance request.
Investing in expensive accelerators to keep them in a standby state is counter-productive. 
It makes more sense to dedicate less devices, accessible from any machine, resulting in a higher utilization rate.
}
\section{\uppercase{GPGPU Shared Service in OpenStack}}
\label{sec:gpgpuOS}
In this section we describe our work on OpenStack, from the development of an external module to handle remote GPUs to the physical setup and its validation.

@@

\subsection{Experimental Results}
The following tests were executed on 2 sets of nodes interconnected by a 1Gb Ethernet network.
All the nodes were equipped with an Intel Xeon E7420 quadcore processor, at 2.13 GHz, and 16 Gbytes of DDR2 RAM memory at 667 MHz.

@@

We have designed 6 different set-ups which can be divided into 2 groups: exclusive and shared GPUs. 
The exclusive mode provides, at most, 4 accelerators. 
The number of available GPU in “shared” mode will depend on the partition size. 
In this particular case, we halved the GPU memory, resulting 8 partitions that can be addressed as independent GPUs.
For each group, we deployed virtual clusters of 1, 2 and 3 nodes, where the application processes were executed. 
The instances were launched with the OpenStack predefined flavor “m1.medium”, which determines a configuration of VMs consisting of 2 cores and 4~GB of memory.

MCUDA-MEME~\cite{Liu2010} was the application selected to test the set-ups. 
Thus is an MPI software, where each process must have access to a GPU.
Therefore, the number of GPUs determines the maximum number of processes we can launch.

Figure~\ref{fig3} compares the execution time of the application with different configurations over different setups. 
We used up to 3 nodes to spread the processes and launched up to 8 processes (only 4 in exclusive mode), one per remote GPU.
We can first observe that the performance is higher with more than one node, because the traffic network is distributed among the nodes.
In addition, the shortest execution time is obtained by both modes (exclusive and shared) when running their maximum number of processes with more than one node.
This seems to imply that it is not worth to scale (increase) the number of resources, because the performance growth rate is barely increasing.
Although, the time is lower when the GPUs are shared, the setup cannot take advantage of an increase in the number of devices.

@@

The most remarkable achievement is the wide range of possible configurations and the flexibility to adapt a system to fit the user requirements.

In addition, with this virtualization technology, the requests for GPU devices can be fulfilled with small investment in infrastructure and maintenance.
Energy can be saved not only thanks to the remote access and the ability to emulate several GPUs using only a few real ones, but also by consolidating the accelerators in a single machine (when possible), or turning down nodes when their GPUs are idle.

@@

\section{\uppercase{Future work}}
\label{sec:future}
The first item in the list of pending work is an upgrade of the network to an interconnect that is more prone to HPC.
In particular, porting the setup and the tests to an infrastructure with an Infiniband network will shed light on the viability of this kind of solutions for HPC.
Similar reasons, motivate us to try other Cloud vendors which better support for HPC.

Looking for real situations where performance is less important than flexibility will drive us to explore alternative tools to easily deploy GPU-programming computer labs.

@@

\section*{\uppercase{Acknowledgements}}
The authors would like to thank the IT members of the department Gustavo Edo and Vicente Roca for their help.
This research was supported by Universitat Jaume~I research project (P11B2013-21); and project
TIN2014-53495-R from MINECO and FEDER.
The initial version of rCUDA was jointly developed by Universitat Polit\`ecnica de 
Val\`encia (UPV) and Universitat Jaume I de Castell\'on (UJI) until year 2010. 
This initial development was later split into two branches. Part of the 
UPV version was used in this paper. The development of the UPV branch 
was supported by Generalitat Valenciana under Grants PROMETEO 2008/060 
and Prometeo II 2013/009. 
\bibliographystyle{apalike}
{\small
\bibliography{paper}}

>>>>

\usepackage{url}% http://ctan.org/pkg/url
%\usepackage{flushend}

@@

\textcolor{black}{
Unfortunately, sharing GPU resources among multiple VMs in cloud environments is more complex than in physical servers.
On one hand, instances in public clouds are not easily customizable. 
On the other, although the instances in a private cloud can be customized in many aspects, when referring to GPUs
the number of options is reduced.
As a result, neither vendors nor tools offer GPGPU services.
}
Remote virtualization has been recently proposed to deal with the low-usage problem. Some relevant examples include {rCUDA}~\cite{tonithesis},
DS-CUDA~\cite{dscuda}, gVirtus~\cite{gvirtus}, 
vCUDA~\cite{vcuda}, VOCL~\cite{vocl}, and SnuCL~\cite{snucl}.
Roughly speaking these virtualization frameworks 
enable cluster configurations with fewer GPUs than nodes.  The goal is that 
GPU-equipped nodes act as GPGPU servers, yielding a GPU-sharing solution that potentially achieves
a higher overall utilization of the accelerators in the system.

@@

The rest of the paper is structured as follows. 
In Section~\ref{sec:background} we introduce the technologies used in this work;
Section~\ref{sec:state} summarizes related work; 
%Section~\ref{sec:motivation} states the motivation to address this task;
the effort to use AWS is explained in Section~\ref{sec:workingaws};
while the work with Openstack is described in Section~\ref{sec:gpgpuOS};
finally, Section~\ref{sec:conclusions} summarizes the advances and Section~\ref{sec:future} outlines the next steps
 of this research.

\section{\uppercase{Background}}
\label{sec:background}
\subsection{The rCUDA Framework}
\label{sec:rcuda}
{rCUDA}~\cite{toniparco} is a middleware that enables transparent access
to any NVIDIA GPU device present in a cluster from all compute
nodes. The GPUs can also be shared among nodes, and a single node can use all the graphic accelerators
as if they were local.
%These features aim to attain higher accelerator utilization rates in the overall system while simultaneously reducing
%resource, space, and energy requirements~\cite{energy14}.
rCUDA is structured following a client-server distributed
architecture and its %: the client middleware runs in the same cluster node where the application demanding GPGPU
%acceleration services is executed, providing a transparent replacement for the
%native CUDA libraries. Furthermore, the server middleware is executed in the
%cluster nodes from which the actual GPUs provide the requested GPGPU services.
%To support a concurrent scenario where GPUs are shared between
%processes\slash nodes, {rCUDA} manages separate device contexts for
%each client application.
%The {rCUDA} 5.0 
client exposes the same interface as the regular NVIDIA
CUDA 6.5 release~\cite{cuda65}.%, including the runtime and driver
%APIs (Application Programming Interface) as well as other frequently used libraries such as cuBLAS, cuFFT, cuSparse or cuRand.
With this middleware, applications are not aware that they are executed
on top of a virtualization layer.
To deal with new GPU programming models, {rCUDA} has been recently extended to accommodate 
directive-based models such as OmpSs~\cite{repara15} and OpenACC~\cite{cluster15}.
The integration of remote GPGPU virtualization with global
resource schedulers such as SLURM~\cite{sbacpad14} completes this appealing 
technology, making accelerator-enabled clusters more flexible and
energy-efficient~\cite{energy14}.

@@

OpenStack~\cite{OpenStack} is a cloud operating system (OS) that provides Infrastructure as a Service (IaaS). 
OpenStack controls large pools of compute, storage, and networking resources throughout a 
datacenter. All these resources are managed through a dashboard or an API that gives administrators 
control while empowering their users to provision resources through a web interface
or a command-line interface.  
OpenStack supports most recent hypervisors and handles provisioning 
and life-cycle management of VMs.
 The OpenStack architecture offers flexibility to create a custom cloud, with no proprietary hardware
or software requirements, and the ability to integrate with legacy systems and third party technologies.

\textcolor{black}{
From the HPC perspective, OpenStack offers high performance virtual machine configurations with different hardware architectures. 
Even though in OpenStack it is possible to work with GPUs, the Nova project does not support this architecture yet.%~\cite{OpenStackGPU}. 
}

@@

\section{\uppercase{GPGPU share service with AWS}}
\label{sec:workingaws}
%In this section, we briefly expose the current user-level features 
%provided by AWS, and then analyze this service  
%from the point of view of GPU usage. 
%Next, we describe the scenarios and experiments where virtualized GPUs are utilized and show a number of initial results.

\subsection{Current Features}
%This subsection identifies several common features checked during 
%the VM's creation step.

@@

An instance is a pre-configured VM focused on an 
specific target. Among the large list of instances offered by AWS, 
we can find specialized versions for general-purpose (T2, M4 and M3); computer science (C4 and C3); memory (R3); storage (I2 and D2) and 
GPU capable (G2). Each type of instance has its own purpose and cost (price). 
Moreover, each type offers a different number of CPUs as well as network 
interconnection, which can be: low, medium, high or 10Gb.
For our study, we worked in the AWS availability zone US EAST (N. VIRGINIA). 
The instances available in that case present the features reported in Table~\ref{table:awsInstances}.

@@

\subsection{Testbed Scenarios}
%In this section, we introduce different client-server configurations. 
All scenarios are based on the maximum number of instances 
that a user can freely select without submitting a formal request. 
In particular, the maximum number for ``g2.2xlarge'' is 5; for ``g2.8xlarge'' it is 2.
Ant the instances operate the RHEL 7.1 64-bit OS and version 6.5 of CUDA. 
We design three configuration scenarios for our tests:

@@

\begin{figure*}[ht]
\centering
\subfigure[5 remote GPUs over 1GbE.]{%
\includegraphics[width=0.3\linewidth]{images/aws3.pdf}
\label{subfig:aws1}}
\quad
\subfigure[8 remote GPUs over 10GbE.]{%
\includegraphics[width=0.3\linewidth]{images/aws2.pdf}
\label{subfig:aws2}}
\subfigure[13 remote GPUs over 1GbE.]{%
\includegraphics[width=0.3\linewidth]{images/aws1.pdf}
\label{subfig:aws3}}
\caption{Configurations where an AWS instance, acting as a client, employs several remote GPUs.}
\label{fig:aws}
\end{figure*}

@@

\subsection{Experimental results}
%In this subsection we employ two GPU applications to assess the benefits of GPU virtualization.

@@

\subsection{Discussion}
\textcolor{black}{The previous experiments reveal that the AWS GPU-instances are not appropriate for HPC
because neither the network nor the accelerators are powerful enough to deliver high performance
when running compute-intensive parallel applications. As~\cite{tonithesis} demonstrates, network and device types are critical factors to performance.
In other words, AWS is more oriented toward offering a general-purpose service than to provide high performance.
Also, AWS fails in offering flexibility as it enforces the user to choose between a basic instance with a GPU and a powerful instance with 4 GPUs.}
Table~\ref{table:awsInstances} shows that the resources of the ``g2.8xlarge'' are quadrupled, but so is the cost per hour.
Therefore, in the case of having other necessities (instances type), using GPU virtualization technology we could in principle attach an accelerator to any type of instance.
Furthermore, reducing the budget spent in cloud services is possible by customizing the resources of the available instances.
For example, we can work on an instance with 2 GPUs for only \$ 1.3 by launching 2 ``g2.2xlarge'' and using remote GPUs, avoiding to pay the double for features that we do not need in ``g2.8xlarge''.
\textcolor{black}{
In terms of GPU-shareability, AWS reserves GPU-capable physical machines which will be waiting for a GPU-instance request.
Investing in expensive accelerators to keep them in a standby state is counter-productive. 
It makes more sense to dedicate less devices, accessible from any machine, resulting in a higher utilization rate.
}
\section{\uppercase{GPGPU Shared Service in OpenStack}}
\label{sec:gpgpuOS}
%In this section we describe our work on OpenStack, from the development of an external module to handle remote GPUs to the physical setup and its validation.

@@

\subsection{Experimental Results}
Following tests were executed on 2 sets of nodes using a 1Gb Ethernet network.
All the nodes composed by an Intel Xeon E7420 quadcore processor, at 2.13 GHz, and 16 GB DDR2 RAM at 667 MHz.

@@

We have designed 6 different set-ups which can be divided into 2 groups: exclusive and shared GPUs. 
The exclusive mode provides, at most, 4 accelerators. 
The number of available GPU in “shared” mode will depend on the partition size. 
In this case, we halved the GPU memory, resulting 8 partitions that can be addressed as independent GPUs.
For each group, we deployed virtual clusters of 1, 2 and 3 nodes, where the application processes were executed. 
The instances were launched with the OpenStack predefined flavor “m1.medium”, which determines a configuration of VMs consisting of 2 cores and 4~GB of memory.

MCUDA-MEME~\cite{Liu2010} was the application selected to test the set-ups. 
Thus is an MPI software, where each process must have access to a GPU.
Therefore, the number of GPUs determines the maximum number of processes we can launch.
Figure~\ref{fig3} compares the execution time of the application with different configurations over different setups. 
We used up to 3 nodes to spread the processes and launched up to 8 processes (only 4 in exclusive mode), one per remote GPU.
We can first observe that the performance is higher with more than one node, because the traffic network is distributed among the nodes.
In addition, the shortest execution time is obtained by both modes (exclusive and shared) when running their maximum number of processes with more than one node.
This seems to imply that it is not worth to scale (increase) the number of resources, because the performance growth rate is barely increasing.
Although, the time is lower when the GPUs are shared, the setup cannot take advantage of an increase in the number of devices.

@@

The most remarkable achievement is the wide range of possible configurations and the flexibility to adapt a system to fit the user requirements. 
In addition, with this virtualization technology, the requests for GPU devices can be fulfilled with small investment in infrastructure and maintenance.
Energy can be saved not only thanks to the remote access and the ability to emulate several GPUs using only a few real ones, but also by consolidating the accelerators in a single machine (when possible), or turning down nodes when their GPUs are idle.

@@

\section{\uppercase{Future work}}
\label{sec:future}
The first item in the list of pending work is an upgrade of the network to an interconnect that is more prone to HPC.
In particular, porting the setup and the tests to an infrastructure with an Infiniband network will shed light on the viability of this kind of solutions.
Similar reasons, motivate us to try other Cloud vendors which better support for HPC.
Looking for situations where performance is less important than flexibility will drive us to explore alternative tools to easily deploy GPU-programming computer labs.

@@

\section*{\uppercase{Acknowledgements}}
The authors would like to thank the IT members of the department Gustavo Edo and Vicente Roca for their help.
This research was supported by Universitat Jaume~I research project (P11B2013-21); and project
TIN2014-53495-R from MINECO and FEDER.
The initial version of rCUDA was jointly developed by Universitat Polit\`ecnica de 
Val\`encia (UPV) and Universitat Jaume I de Castell\'on until year 2010. 
This initial development was later split into two branches. Part of the 
UPV version was used in this paper and it  
was supported by Generalitat Valenciana under Grants PROMETEO 2008/060 
and Prometeo II 2013/009. 
\bibliographystyle{apalike}
{\small
\bibliography{paper}}
